Multivariate Density Estimation: Comparing Transformation Random Forest, Normalizing Flows and Copulas

### 2025-09-30 report/MSc_Report.Rnw
.

\chapter*{Abstract}
% Checklist (Abstract): [x] integrate supplied summary [x] confirm PDF rebuild

This thesis evaluates three approaches to multivariate density estimation for tabular data within a single, consistent pipeline: separable triangular transport maps (TTM-Sep), Transformation Random Forests (TRTF), and copulas. All methods use standardized inputs and a common evaluation protocol so that likelihoods, diagnostics, and compute are directly comparable. In the configuration studied, TRTF and TTM-Sep yield the same triangular-likelihood form, which enables like-for-like evaluation.

On Half-Moon ($n=250$), mean joint negative log-likelihoods (NLL; lower is better) were $1.71$ (TRTF), $1.93$ (TTM-Sep), and $1.54$ (copula). On a four-dimensional autoregressive generator they were $4.53$, $5.66$, and $5.45$, respectively; permutation averages confirm order sensitivity for triangular maps. On MiniBooNE ($K=43$; sum test log-likelihood), TRTF reached $-30.01$ under the standard preprocessing and training budget used here; published flow models report values around $-12$ to $-16$ under their settings. These numbers are not strictly comparable but indicate the relative accuracy of this configuration.

>>>>

\chapter*{Abstract}

This thesis evaluates three approaches to multivariate density estimation for tabular data within a single, consistent pipeline: separable triangular transport maps (TTM-Sep), Transformation Random Forests (TRTF), and copulas. All methods use standardized inputs and a common evaluation protocol so that likelihoods and diagnostics are directly comparable; compute summaries are reported for TRTF (Table~3.7). In the configuration studied, TRTF and TTM-Sep yield the same triangular-likelihood form, which enables like-for-like evaluation. Separable triangular maps deliver exact likelihoods, O(K) evaluation/inversion, and transparent conditionals, but restrict context-dependent shape.

On Half-Moon ($n=250$), mean joint negative log-likelihoods (NLL; lower is better) were $1.71$ (TRTF), $1.93$ (TTM-Sep), and $1.54$ (copula). On a four-dimensional autoregressive generator they were $4.53$, $5.66$, and $5.45$, respectively; permutation averages confirm order sensitivity for triangular maps. On MiniBooNE ($K=43$; mean test log-likelihood per sample), TRTF reached $-30.01$ under the standard preprocessing and training budget used here; published flow models report values around $-12$ to $-16$ under their settings. These numbers are not strictly comparable but indicate the relative accuracy of this configuration.
